Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-02-17 draft.tex
Updates from Overleaf

Type Ia supernovae (SNe Ia) are explosions of white dwarf stars (WDs) at masses close to the Chandrasekhar limit of 1.44 $M_{\odot}$. The small observed dispersion in their peak brightnesses can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour \citep{Philips1993,Tripp1998}. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe \citep{Perlmutter1998,Riess1998}

@@

In this work, we combine parts of the above methods and derive a new, independent measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012}. We use the stellar mass assembly model of \citet{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit by a two-component DTD.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. Such a distinction between SN stretch and progenitor age has been seen by \citet{Rigault2013,Childress2014,Rigault2018,Nicolas2020}; here we have shown it purely by measuring the rate against stellar mass.

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would both follow the fiducial $t^{-1}$ distribution.
Such a scenario has been explored theoretically, primarily in the sub-Chandrasekhar mass (sub-$M_{\mathrm{Ch}}$) paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}, usually associated with various forms of the DD scenario. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs where the SN luminosity is correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation.

>>>>

Type Ia supernovae (SNe Ia) are explosions of white dwarf stars (WDs) at masses close to the Chandrasekhar limit of 1.44 $M_{\odot}$ ($M_{\mathrm{Ch}}$. The small observed dispersion in their peak brightnesses can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour \citep{Phillips1993,Tripp1998}. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe \citep{Perlmutter1998,Riess1998}

@@

While the majority of observations based on populations point towards a DD scenario for most, if not all, SNe Ia, finding self-consistent explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs fail to reproduced the observed relationship between SN luminosity and light curve duration (the Phillips relation; \citealt{Phillips1993}) as well as low-luminosity "peculiar" objects known as SN 1991bg-like SNe \citep{Sim2013,Blondin2017}. In recent years, attention focussed on explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations has led to promising results \citep[e.g.][]{Chen2017,Chen2018,Townsley2019,Gronow2020}. An additional factor in support of the double-detonation model is that the SN luminosity is related to the mass of the primary WD which can be related to its age, thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Nicolas2020}

In this work, we combine parts of the traditional methods and derive a new, independent measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012}. We use the stellar mass assembly model of \citet{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit by a two-component DTD.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Rigault2018,Nicolas2020}. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population.

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would both follow the fiducial $t^{-1}$ distribution.
Such a scenario may be explained by the sub-$M_{\mathrm{Ch}}$ double-detonation paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017a}. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs and the SN luminosity can be correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation.
